---
title: "DS203-Lab2"
output: pdf_document
date: "2024-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data <- read.csv("Downloads/final_data.csv")
head(data)

```

First create dummy variable for both Gender and Poverty_Index

```{r}
# Create dummy variables for Gender and Poverty_Index
data$Gender_Male <- ifelse(data$RIAGENDR == 1, 1, 0)
# male is 1, female is 2 so coded it to be male is 1, female is 0
data$Poverty_Index_Binary <- ifelse(data$Poverty_Index > 2.5, 1, 0)
# the index is above 2.5, then encode as 1. Otherwise, 0. 
```
# Simple Model - Variable Analysis
First let's just assess the variables Age vs. Sleep. 

```{r}

# age histogram

age_hist <- ggplot(data, aes(x = Age)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black", alpha = 0.7) +
  ggtitle("Histogram of Age") +
  xlab("Age") +
  ylab("Frequency")

age_hist

# sleep histogram 

sleep_hist <- ggplot(data, aes(x = Sleep_Hours_Weekday)) +
  geom_histogram(bins = 20, fill = "lightgreen", color = "black", alpha = 0.7) +
  ggtitle("Histogram of Sleep Hours") +
  xlab("Sleep Hours") +
  ylab("Frequency")

sleep_hist

# joint relationship of age vs sleep

agesleep_plot <- ggplot(data, aes(x = Age, y = Sleep_Hours_Weekday)) +
  geom_point(alpha = 0.5) +
  ggtitle("Age vs Sleep Hours") +
  xlab("Age") +
  ylab("Sleep Hours")

agesleep_plot

```
 Let's perform a quadratic transformation to see if we can account for a non-linear relationship between Age and Sleep. 

```{r}
# Create a quadratic term for Age
data$Age_Squared <- data$Age^2

# Fit a linear regression model (Age as a predictor)
model_linear_age <- lm(Sleep_Hours_Weekday ~ Age, data = data)

# Fit a quadratic regression model (Age and Age^2 as predictors)
model_quadratic_age <- lm(Sleep_Hours_Weekday ~ Age + Age_Squared, data = data)

# Compare the models
summary(model_linear_age)
summary(model_quadratic_age)
```

While the quadratic model is statistically better, the overall explanatory power (2.1%) remains very low. This suggests that Age (even with non-linear terms) has a small impact on Sleep Hours.


Visualize the relationship between Age and Sleep (linear and quadratic models)
```{r}
# Predictions from both models
data$Predicted_Linear <- predict(model_linear_age, data)
data$Predicted_Quadratic <- predict(model_quadratic_age, data)

# Plot the actual data points and fitted lines
library(ggplot2)

ggplot(data, aes(x = Age, y = Sleep_Hours_Weekday)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = Predicted_Linear), color = "blue", linetype = "dashed") +
  geom_line(aes(y = Predicted_Quadratic), color = "red") +
  ggtitle("Linear vs Quadratic Model: Age vs Sleep Hours") +
  xlab("Age") +
  ylab("Sleep Hours") +
  theme_minimal()
```
quadratic model better fits the data but has limited explanatory power.



let's evaluate Gender Variable: 
Since Gender is a categorical variable (1 = Male, 0 = Female), use boxplots and summary statistics to explore differences in Sleep Hours.

```{r}
# Visualize the relationship using a boxplot
library(ggplot2)
ggplot(data, aes(x = factor(Gender_Male), y = Sleep_Hours_Weekday)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  ggtitle("Boxplot of Gender vs Sleep Hours") +
  xlab("Gender (0 = Female, 1 = Male)") +
  ylab("Sleep Hours") +
  theme_minimal()

# Calculate summary statistics for each gender
aggregate(Sleep_Hours_Weekday ~ Gender_Male, data = data, summary)
```
```{r}
# Fit a linear model with Gender as the predictor
model_gender <- lm(Sleep_Hours_Weekday ~ Gender_Male, data = data)

# View the model summary
summary(model_gender)
```
The median sleep hours for Males (Gender = 1) and Females (Gender = 0) appear very close, suggesting minimal differences.
The interquartile range (IQR) and spread of the data are also similar between the two genders.
There are no dramatic outliers specific to one gender, indicating that extreme sleep values are equally distributed.
overall model is not statistically significant, further confirming that Gender is not a strong predictor of Sleep Hours.

Analyze Poverty Index vs. Sleep. We will be treating this as a dummy variable with > 2.5 as High Poverty Index and less than that as low with 0 = Low and 1 = High. 

```{r}
# Visualize the relationship using a boxplot
ggplot(data, aes(x = factor(Poverty_Index_Binary), y = Sleep_Hours_Weekday)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  ggtitle("Boxplot of Poverty Index vs Sleep Hours") +
  xlab("Poverty Index (0 = Low, 1 = High)") +
  ylab("Sleep Hours") +
  theme_minimal()

# Calculate summary statistics for each group
aggregate(Sleep_Hours_Weekday ~ Poverty_Index_Binary, data = data, summary)

```
```{r}
# Fit a linear model with Poverty Index as the predictor
model_poverty <- lm(Sleep_Hours_Weekday ~ Poverty_Index_Binary, data = data)

# View the model summary
summary(model_poverty)
```

The Adjusted R-squared is extremely low, indicating that the Poverty Index does not meaningfully explain variance in Sleep Hours on its own.
Include Poverty Index: The statistically significant p-value justifies including this variable in the overall model, especially since it could interact with other predictors (e.g., Age or Gender).
However, its practical impact is minimal, so you should emphasize this in your report.

Now fit the full model (model 1 with 3 features)
```{r}
# Fit the multiple linear regression model
model_full <- lm(Sleep_Hours_Weekday ~ Age + Age_Squared + Gender_Male + Poverty_Index_Binary, data = data)

# View the summary of the model
summary(model_full)

```

# Simple model - Assumptions

1. IID Assumption 
Plot Residuals vs Fitted Values, residuals should show no patterns and should be randomly distributed. If we see patterns, the IID assumption may be violated
```{r}
# Plot Residuals vs Fitted
plot(model_full, which = 1)  # Residuals vs Fitted

```

```{r}
# Residual diagnostics
plot(model_full, which = 1)  # Residuals vs Fitted
plot(model_full, which = 2)  # Q-Q Plot
```

```{r}
# Breusch-Pagan test for heteroskedasticity
library(lmtest)
bptest(model_full)
```

```{r}
# Shapiro-Wilk test for normality of residuals
shapiro.test(residuals(model_full))
```


```{r}
# Check for multicollinearity using VIF
install.packages("car")
library(car)
vif(model_full)
```

# Complex Model - Adding LTPA
We will now plot the relationship for Total_Week_LTPA vs Sleep
```{r}
# Scatter plot for Total_Week_LTPA vs Sleep Hours
library(ggplot2)
ggplot(data, aes(x = Total_Week_LTPA, y = Sleep_Hours_Weekday)) +
  geom_point(alpha = 0.5, color = "blue") +
  ggtitle("Total_Week_LTPA vs Sleep Hours") +
  xlab("Total_Week_LTPA") +
  ylab("Sleep Hours") +
  theme_minimal()

```
Since Total_Week_LTPA is highly skewed, apply a log transformation to stabilize the variance and make patterns more interpretable.
```{r}
# Add a log-transformed version of Total_Week_LTPA (add 1 to avoid log(0))
data$Log_Total_Week_LTPA <- log(data$Total_Week_LTPA + 1)

# Scatter plot with log-transformed Total_Week_LTPA
ggplot(data, aes(x = Log_Total_Week_LTPA, y = Sleep_Hours_Weekday)) +
  geom_point(alpha = 0.5, color = "green") +
  ggtitle("Log-Transformed Total_Week_LTPA vs Sleep Hours") +
  xlab("Log_Total_Week_LTPA") +
  ylab("Sleep Hours") +
  theme_minimal()

```

```{r}
# Fit the full model including Log_Total_Week_LTPA
model_full_with_ltp <- lm(Sleep_Hours_Weekday ~ Age + Age_Squared + Gender_Male + 
                          Poverty_Index_Binary + Log_Total_Week_LTPA, data = data)

# View the model summary
summary(model_full_with_ltp)

```
```{r}
install.packages("stargazer")
library(stargazer)
stargazer(model_full_with_ltp, type = "text")

```

Model Fit:
Adjusted R-squared: 0.02216
The model explains only 2.2% of the variance in Sleep Hours. While statistically significant (p < 0.001), the model has very low explanatory power


# ASSUMPTIONS for COMPLEX MODEL; 
```{r}
# Residual diagnostics
plot(model_full_with_ltp, which = 1)  # Residuals vs Fitted
plot(model_full_with_ltp, which = 2)  # Q-Q Plot
```

```{r}
# Breusch-Pagan test for heteroskedasticity
library(lmtest)
bptest(model_full_with_ltp)
```

```{r}
library(car)
vif(model_full_with_ltp)
```

